Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change

### 2021-01-21 Manuscript/HI_Model_ms.Rmd
updated manuscript files

The State of Hawaiʻi exemplifies the challenges of projecting the interactive effects of future climate and land use change on ecosystem carbon balance at a regional scale. Hawaiʻi was the first U.S. state to enact legislation committing to full carbon neutrality, requiring the state to account for and offset all of its greenhouse gas emissions by 2045 (State of Hawaiʻi Acts 15 and 16). This legislation emphasizes the mitigation potential of natural ecosystems as a key component to emissions reduction, necessitating baseline estimates and future projections of land carbon sink strength. However, Hawaiʻi's challenging terrain complicates these assessment efforts. The main Hawaiian Islands are a complex mosaic of natural and human-dominated landscapes overlain by steep climate gradients across relatively short distances, with mean annual temperature ranging from ~4-24° C [@giambelluca_evapotranspiration_2014] and mean annual rainfall ranging from ~180-9500 mm [@giambelluca_online_2013]. Temperatures have risen rapidly in the Hawaiian Islands since the mid 1970s [@giambelluca_secular_2008] and a long-term drying trend has persisted since the early 1920s [@frazier_spatial_2017], resulting in reduced forest biomass and productivity [@barbosa_effects_2017]. These same drying and warming trends have increased the frequency and intensity of wildland fire [@trauernicht_contemporary_2015; @trauernicht_vegetationrainfall_2019] with predictable negative effects on ecosystem carbon balance [@selmants_baseline_2017]. Ecosystem carbon stocks across the main Hawaiian Islands have also been strongly influenced by the legacy of past land-use change [@osher_effect_2003; @asner_high-resolution_2011]. Thousands of hectares of land were deforested beginning in the late 19th century to clear land for sugar plantations and cattle pasture [@cuddihy_alteration_1990]. Since the mid-20th century, much of this agricultural land has been steadily converted to urban areas, commercial forestry plantations, or simply abandoned and colonized by non-native grass species [@suryanata_diversified_2009; @perroy_evolving_2016].

>>>>

The State of Hawaiʻi exemplifies the challenges associated with projecting the interactive effects of future climate and land use change on ecosystem carbon balance at a regional scale. Hawaiʻi was the first U.S. state to enact legislation committing to full carbon neutrality, requiring the state to account for and offset all of its greenhouse gas emissions by 2045 (State of Hawaiʻi Acts 15 and 16). This legislation emphasizes the mitigation potential of natural ecosystems as a key component to emissions reduction, necessitating baseline estimates and future projections of land carbon sink strength. However, Hawaiʻi's challenging terrain complicates these assessment efforts. The main Hawaiian Islands are a complex mosaic of natural and human-dominated landscapes overlain by steep climate gradients across relatively short distances, with mean annual temperature ranging from ~4-24° C [@giambelluca_evapotranspiration_2014] and mean annual rainfall ranging from ~180-9500 mm [@giambelluca_online_2013]. Temperatures have risen rapidly in the Hawaiian Islands since the mid 1970s [@giambelluca_secular_2008] and a long-term drying trend has persisted since the early 1920s [@frazier_spatial_2017], resulting in reduced forest biomass and productivity [@barbosa_effects_2017]. These same drying and warming trends have increased the frequency and intensity of wildland fire [@trauernicht_contemporary_2015; @trauernicht_vegetationrainfall_2019] with predictable negative effects on ecosystem carbon balance [@selmants_baseline_2017]. Ecosystem carbon stocks across the main Hawaiian Islands have also been strongly influenced by the legacy of past land-use change [@osher_effect_2003; @asner_high-resolution_2011]. Thousands of hectares of land were deforested beginning in the late 19th century to clear land for sugar plantations and cattle pasture [@cuddihy_alteration_1990]. Since the mid-20th century, much of this agricultural land has been steadily converted to urban areas, commercial forestry plantations, or simply abandoned and colonized by non-native grass species [@suryanata_diversified_2009; @perroy_evolving_2016]. These past trends surely inform the future impact of climate and land-use change on ecosystem carbon balance, but remain difficult to predict with any degree of certainty. To date only one study has attempted to integrate the major controlling processes of land-use, climate, and natural disturbances into projections of carbon balance for Hawaiian ecosystems, and that study was restricted to a single land-use change scenario, a single radiative forcing scenario (SRES A1B, equivalent to RCP 6), and a projection period extending only to the mid-21st century [@selmants_baseline_2017].
